Self-assembly of colloid-cholesteric composites provides a possible route to switchable optical materials

### 2014-04-17 trunk/papers/2012-sandwich/CholColloids.tex
Further minor adjustments
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2364 40876441-3496-0410-a613-a2fca35b2ae2

Besides being remarkable materials in themselves, BPs offer significant promise as hosts for dispersing colloidal particles. Because BPs contain a disclination network even in the absence of
particles, they can potentially template colloidal self-assembly. 
This was proved recently by simulations of BP-dispersed nanoparticles which 
exclude the surrounding liquid crystal but otherwise have negligible surface 
interaction with it (to give a so-called `weak anchoring' regime)~\cite{extrareference1,miha}. 
Such particles are attracted to the disclination network in BPs: by covering up the defect cores, the high elastic energy cost of those regions is avoided. The pre-existing order of the defect network then templates particles into a regular colloidal crystal of the same periodicity. 
Because this structure has a wavelength in the visible range, 
the resulting material should inherit (and likely enhance) the incomplete photonic bandgap of the parent blue phase, as exploited for instance in laser devices~\cite{bplasers}; for related possibilities see \cite{lavrentovich}.

@@

\bibitem{bp3} Henrich, O., Stratford, K., Cates, M.E., and Marenduzzo, D.,
 Structure of blue phase III of cholesteric liquid crystals,
{\it Phys. Rev. Lett.} {\bf 106}, 107802 (2011).

@@

\bibitem{bplasers}
Cao, W.Y., Munoz, A., Palffy-Muhoray, P., and  Taheri, B.,
Lasing in a three-dimensional photonic crystal of the liquid crystal blue 
phase II,
{\it Nat. Mat.} {\bf 1}, 111--113 (2002).

>>>>

Besides being remarkable materials in themselves, BPs offer significant promise as hosts for dispersing colloidal particles. Because BPs contain a disclination network even in the absence of
particles, they can potentially template colloidal self-assembly. 
This was proved recently by simulations of BP-dispersed nanoparticles which 
exclude the surrounding liquid crystal but otherwise have negligible surface 
interaction with it (to give a so-called `weak anchoring' regime)~\cite{extrareference1,miha}. 
Such particles are attracted to the disclination network in BPs: by covering up the defect cores, the high elastic energy cost of those regions is avoided. The pre-existing order of the defect network then templates particles into a regular colloidal crystal of the same periodicity. 
Because this structure has a wavelength in the visible range, 
the resulting material should inherit (and probably enhance) the incomplete photonic bandgap of the parent blue phase, as exploited for instance in laser devices~\cite{bplasers}; for related possibilities see \cite{lavrentovich}.

@@

\bibitem{bp3} Henrich, O., Stratford, K., Cates, M.E., and Marenduzzo, D.,
 Structure of blue phase~III of cholesteric liquid crystals,
{\it Phys. Rev. Lett.} {\bf 106}, 107802 (2011).

@@

\bibitem{bplasers}
Cao, W.Y., Munoz, A., Palffy-Muhoray, P., and  Taheri, B.,
Lasing in a three-dimensional photonic crystal of the liquid crystal blue 
phase~II,
{\it Nat. Mat.} {\bf 1}, 111--113 (2002).

### 2014-04-17 trunk/papers/2012-sandwich/support.tex
Further minor adjustments
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2364 40876441-3496-0410-a613-a2fca35b2ae2

{\bf
Self-assembly of colloid-cholesteric
composites provides a possible route to switchable optical materials.

Supplementary Information}

Here, we give more details of the simulation method, protocol, and parameters used. We also provide additional figures complementing the main text, and discuss the matching of simulation units to physical units.

@@

The surface anchoring for colloids is as before: always normal, but with
$WR/K = 0.23$ and $WR/K = 2.3$. The colloid solid volume fractions are
1\% and 4\%.
\textcolor{red}{For both normal and planar anchoring at the walls,
the strength is adjusted to correspond to that used for the strongly
anchoring particles.}
All simulations are run for two million simulation time steps.

\noindent
\textbf{External electric field.}
For the simulations of BPIII with colloidal particles a system size of
$128^3$ is used with a cholesteric pitch of $p=32$ in lattice units. 
The simulations are
performed at chirality $\kappa=2.6$ and reduced temperature $\tau=-0.25$.  
An initial configuration of
randomly oriented and positioned double twist cylinders is used,
from which an amorphous BPIII network emerges \cite{oliver-bp3}. This structure
is equilibrated for $6\times10^{5}$ LB time steps until no significant
further evolution is observed.

@@

\begin{table}[h]
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|c|c|}
\hline
         & $A_0$ & $\gamma$ & $K$ & $p$ & $\kappa$ & $\tau$ & $WR/K$\\
\hline 
Bulk BPI & 0.01 & 3.086 & 0.007061 & 32$\sqrt{2}$ & 0.6902 & -0.2500 &
0.23, 2.3\\
\hline
Bulk Quench & 0.01 & 3.086 & 0.007061 & 32$\sqrt{2}$ & 0.6902 & -0.2500 &
0.23, 2.3 \\
\hline
Confined & 0.01 & 3.086 & 0.01897 & 64 & 0.8 & -0.2500 & 0.23, 2.3\\
\hline
External field & 0.004& 3.086 & 0.02 & 32 & 2.6 & -0.2500 & 0.23\\ 
\hline
\end{tabular}
\end{center}
\caption{\textbf{Free energy parameters used in the simulations.} Bulk BPI
correspond to Fig.~1 (A-D), bulk quench to Fig.~1 (E--H); confined
geometries are shown in Fig.~2; and the external field case is
relevant for Fig.~3.}
\label{table:params}
\end{table}

\noindent
\textbf{Disclination rendering.}
In all cases visualisation of the defect lines is carried out via
an isosurface of the scalar order parameter $q$ determined from the
largest eigenvalue of the local order parameter tensor. A low value
of the scalar order parameter $q$ (typically in the range 0.12--0.14)
unambiguously identifies the disclinations.

@@

The timescale calibration may be obtained from (see e.g.~\cite{denniston})
\begin{equation}
\gamma_1=\frac{2q^2}{\Gamma},
\end{equation}
which relates the rotational viscosity $\gamma_1$ to the ordering strength
$q$ and the order parameter mobility $\Gamma$.
In the present simulations, $\Gamma = 0.5$ and the value of $\gamma$
for the simulation in Fig.~1 leads to $q=1/2$.   
For real liquid crystalline materials, $\gamma_1$ usually lies in range
$10^{-2}-1$ Pa s~\cite{deGennes}; for definiteness say
$\gamma_1 = 1$~Pa~s (equivalently 10 poise). Given the previous mapping for
free energy (or stress) units, this leads to a simulation unit of time equal
to $10^{-8}$~s. The total simulated time in Figs. 1-3 is therefore in the
millisecond range.

To calibrate the electric field strength 
one can equate the value of the dimensionless parameter ${\cal E}$ in the 
simulations with that of a hypothetical experiment.
Given the previous value for $A_0$, and a dielectric anisotropy of 20
(or equivalently $\epsilon_a=20\epsilon_0$, with $\epsilon_0$ the
dielectric permittivity of vacuum), an electric field
of 100 V~$\mu$m$^{-1}$ is obtained. This corresponds to a value of
${\cal E}\sim 0.4$ (considering a value of $\gamma=3.0857$ as in
Fig.~3 of the main text).

@@

To summarise, the simulations represent BP-forming materials, with the
interpretation of the simulation units for length, time, and energy
density being close to 10~nm, 10~ns, and 100~MPa respectively.
For a Frank eleastic
constant of $K = 70$~pN and unit cell sizes in the range
$\lambda =$100--500~nm, a corresponding shear modulus $G \simeq
K/\lambda^2 =$0.3--7~kPa.

{\bf Supplementary Figures and Animations}

\begin{figure}[!h]
\begin{center}
\includegraphics[scale=0.35]{support-fig2.pdf}
\end{center}
\caption{\textbf{Bulk blue phase structures.}
As for main manuscript Fig.~1 A--D
(with liquid crystal order parameter
initialised to equilibrium blue phase I structure), but for the entire
simulation system of 128$^3$ lattice sites. (A)  solid volume fraction~1\%
and weak anchoring;  (B) 4\% solid volume fraction and weak anchoring;
(C) 1\% solid volume fraction and strong anchoring; (D) 4\% solid
volume fraction and strong anchoring. The view direction in the
main figure is from the left here. There is a reference
particle at the bottom left in each panel which does not take part
in the simulation.}
\end{figure}

\begin{figure}[!h]
\begin{center}
\includegraphics[scale=0.35]{support-fig3.pdf}
\end{center}
\caption{\textbf{Bulk blue phase structures.}
As for main manuscript Fig.~1 E--H (initialised via a ``quench''
to generate a disordered network), but for the entire simulation
system of 128$^3$ lattice sites. (E) solid volume fraction 1\% and
weak anchoring; (F) 4\% solid volume fraction and weak anchoring;
(G) 1\% solid volume fraction and strong anchoring; (H) 4\% solid
volume fraction and strong anchoring. Again, there is a reference
particle at the bottom left in each case which does not take part
in the simulation.}
\end{figure}

\begin{figure}[!h]
\begin{center}
\includegraphics[scale=0.49]{support-fig4.png}
\end{center}
\caption{\textbf{Free energy evolution.}
Plot of the free energy (per BP unit cell) versus time
for the simulations corresponding to Fig. 1A-H in the main text
(the labels correspond to those in Fig. 1).
}
\end{figure}

\begin{figure}[!h]
\begin{center}
\includegraphics[scale=0.42]{support-fig5.pdf}
\end{center}
\caption{\textbf{Confined blue phase structures.}
As for main manuscript Fig.~2 A--D (confined geometry with normal
anchoring walls at both top and bottom), but for the entire simulation
system of 256$^2\times$56 lattice sites. Each shows a top view and
side view of the same simulation state. (A) solid volume fraction 1\%
and weak anchoring; (B) 4\% solid volume fraction and weak anchoring;
(C) 1\% solid volume fraction and strong anchoring; (D) 4\% solid
volume fraction and strong anchoring.}
\end{figure}

\begin{figure}[!h]
\begin{center}
\includegraphics[scale=0.42]{support-fig6.pdf}
\end{center}
\caption{\textbf{Confined blue phase structures.}
As for main manuscript Fig.~2 E--H (confined geometry with planar
anchoring walls at both top and bottom), but for the entire simulation
system of 256$^2\times$56 lattice sites. Each shows a top view and
side view of the same simulation state. (E) colloid solid volume fraction 1\%
and weak anchoring; (F) 4\% solid volume fraction and weak anchoring;
(G) 1\% solid volume fraction and strong anchoring; (H) 4\% solid
volume fraction and strong anchoring.}
\end{figure}

\begin{figure}[!h]
\begin{center}
\includegraphics[width=0.34\columnwidth]{col_parallel_x_run1341.png}
\includegraphics[width=0.32\columnwidth]{sq_x_run1341_2.png}
\includegraphics[width=0.32\columnwidth]{sq_y_run1341.png}
\includegraphics[width=0.34\columnwidth]{col_parallel_x_run1342.png}
\includegraphics[width=0.32\columnwidth]{sq_x_run1342.png}
\includegraphics[width=0.32\columnwidth]{sq_y_run1342.png}
\end{center}
\caption{\textbf{Structure with and without electric field.}
(Disclination lines are 
removed for clarity.) Panels A and B reproduce the results in Fig. 3C and
Fig. 3D respectively with the applied field in the $x$-direction 
(out of the plane of the paper). Also included is a cut of the structure 
factor with $k_y = 0$ (panel C). 
Panels D, E, and F show the corresponding situation after the
external field has been removed and the structure allowed to relax,
and show residual hexagonal ordering. The structure factor data is for wave
vectors $k_x = 0; (k_y, k_z) \in [-3\pi/8 , 3\pi/8]$ (panels B and E) and
$k_y = 0; (k_x, k_z) \in [-3\pi/8, 3\pi/8]$ (panels C and F).}
\end{figure}

\begin{figure}[!h]
\begin{center}
\includegraphics[width=0.34\columnwidth]{col_parallel_y_run1343.png}
\includegraphics[width=0.32\columnwidth]{sq_x_run1343.png}
\includegraphics[width=0.32\columnwidth]{sq_y_run1343.png}\\
\includegraphics[width=0.34\columnwidth]{col_parallel_y_run1344.png}
\includegraphics[width=0.32\columnwidth]{sq_x_run1344.png}
\includegraphics[width=0.32\columnwidth]{sq_y_run1344.png}\\
\end{center}
\caption{\textbf{Cycling the electric field.}
Starting from the final position of
Supplementary Fig.~7 panel~D, the field is re-applied, this time in the
$y$-direction. The corresponding view of the particle distribution is
shown, again viewing along the field direction, in panel A.
Corresponding structure factor cuts with $k_x = 0$ and $k_y = 0$ are
shown in panels B and C, respectively. The corresponding situation when
the field is again switched off, and the structure allowed to relax,
is shown in panels D, E, and F. The wave vector values used are the
same as in Supplementary Fig.~7.}
\end{figure}

>>>>

{\bf Supplementary Figures}

\begin{figure}[!h]
\begin{center}
\includegraphics[scale=0.35]{support-fig2.pdf}
\end{center}
\caption{\textbf{Bulk blue phase structures.}
As for main manuscript Fig.~1~\textbf{a}--\textbf{d}
(with liquid crystal order parameter
initialised to equilibrium blue phase~I structure), but for the entire
simulation system of 128$^3$ lattice sites. (A)~solid volume fraction~1\%
and weak anchoring;  (B)~4\% solid volume fraction and weak anchoring;
(C) 1\% solid volume fraction and strong anchoring; (D) 4\% solid
volume fraction and strong anchoring. The view direction in the
main figure is from the left here. There is a reference
particle at the bottom left in each panel which does not take part
in the simulation.}
\end{figure}

\begin{figure}[!h]
\begin{center}
\includegraphics[scale=0.35]{support-fig3.pdf}
\end{center}
\caption{\textbf{Bulk blue phase structures.}
As for main manuscript Fig.~1 \textbf{e}--\textbf{h}
(initialised via a ``quench''
to generate a disordered network), but for the entire simulation
system of 128$^3$ lattice sites. (E) solid volume fraction 1\% and
weak anchoring; (F) 4\% solid volume fraction and weak anchoring;
(G) 1\% solid volume fraction and strong anchoring; (H) 4\% solid
volume fraction and strong anchoring. Again, there is a reference
particle at the bottom left in each case which does not take part
in the simulation.}
\end{figure}

\begin{figure}[!h]
\begin{center}
\includegraphics[scale=0.49]{support-fig4.png}
\end{center}
\caption{\textbf{Free energy evolution.}
Plot of the free energy (per BP unit cell) versus time
for the simulations corresponding to Fig.~1 (\textbf{a}--\textbf{h})
in the main text.}
\end{figure}

\begin{figure}[!h]
\begin{center}
\includegraphics[scale=0.42]{support-fig5.pdf}
\end{center}
\caption{\textbf{Confined blue phase structures.}
As for main manuscript Fig.~2 \textbf{a}--\textbf{d}
(confined geometry with normal
anchoring walls at both top and bottom), but for the entire simulation
system of 256$^2\times$56 lattice sites. Each shows a top view and
side view of the same simulation state. (A) solid volume fraction 1\%
and weak anchoring; (B) 4\% solid volume fraction and weak anchoring;
(C) 1\% solid volume fraction and strong anchoring; (D) 4\% solid
volume fraction and strong anchoring.}
\end{figure}

\begin{figure}[!h]
\begin{center}
\includegraphics[scale=0.42]{support-fig6.pdf}
\end{center}
\caption{\textbf{Confined blue phase structures.}
As for main manuscript Fig.~2 \textbf{e}--\textbf{h}
(confined geometry with planar
anchoring walls at both top and bottom), but for the entire simulation
system of 256$^2\times$56 lattice sites. Each shows a top view and
side view of the same simulation state. (E) colloid solid volume fraction 1\%
and weak anchoring; (F) 4\% solid volume fraction and weak anchoring;
(G) 1\% solid volume fraction and strong anchoring; (H) 4\% solid
volume fraction and strong anchoring.}
\end{figure}

\begin{figure}[!h]
\begin{center}
\includegraphics[width=0.34\columnwidth]{col_parallel_x_run1341.png}
\includegraphics[width=0.32\columnwidth]{sq_x_run1341_2.png}
\includegraphics[width=0.32\columnwidth]{sq_y_run1341.png}
\includegraphics[width=0.34\columnwidth]{col_parallel_x_run1342.png}
\includegraphics[width=0.32\columnwidth]{sq_x_run1342.png}
\includegraphics[width=0.32\columnwidth]{sq_y_run1342.png}
\end{center}
\caption{\textbf{Structure with and without electric field.}
(Disclination lines are 
removed for clarity.) Panels~A and~B reproduce the results in Fig.~3\textbf{c}
and
Fig.~3\textbf{d} respectively with the applied field in the $x$-direction 
(out of the plane of the paper). Also included is a cut of the structure 
factor with $k_y = 0$ (panel C). 
Panels D, E, and F show the corresponding situation after the
external field has been removed and the structure allowed to relax,
and show residual hexagonal ordering. The structure factor data is for wave
vectors $k_x = 0; (k_y, k_z) \in [-3\pi/8 , 3\pi/8]$ (panels B and E) and
$k_y = 0; (k_x, k_z) \in [-3\pi/8, 3\pi/8]$ (panels C and F).}
\end{figure}

\begin{figure}[!h]
\begin{center}
\includegraphics[width=0.34\columnwidth]{col_parallel_y_run1343.png}
\includegraphics[width=0.32\columnwidth]{sq_x_run1343.png}
\includegraphics[width=0.32\columnwidth]{sq_y_run1343.png}\\
\includegraphics[width=0.34\columnwidth]{col_parallel_y_run1344.png}
\includegraphics[width=0.32\columnwidth]{sq_x_run1344.png}
\includegraphics[width=0.32\columnwidth]{sq_y_run1344.png}\\
\end{center}
\caption{\textbf{Cycling the electric field.}
Starting from the final position of
Supplementary Fig.~7 panel~D, the field is re-applied, this time in the
$y$-direction. The corresponding view of the particle distribution is
shown, again viewing along the field direction, in panel A.
Corresponding structure factor cuts with $k_x = 0$ and $k_y = 0$ are
shown in panels~B and~C, respectively. The corresponding situation when
the field is again switched off, and the structure allowed to relax,
is shown in panels D, E, and~F. The wave vector values used are the
same as in Supplementary Fig.~7.}
\end{figure}

{\bf Supplementary Table}

\begin{table}[h]
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|c|c|}
\hline
         & $A_0$ & $\gamma$ & $K$ & $p$ & $\kappa$ & $\tau$ & $WR/K$\\
\hline 
Bulk BPI & 0.01 & 3.086 & 0.007061 & 32$\sqrt{2}$ & 0.6902 & -0.2500 &
0.23, 2.3\\
\hline
Bulk Quench & 0.01 & 3.086 & 0.007061 & 32$\sqrt{2}$ & 0.6902 & -0.2500 &
0.23, 2.3 \\
\hline
Confined & 0.01 & 3.086 & 0.01897 & 64 & 0.8 & -0.2500 & 0.23, 2.3\\
\hline
External field & 0.004& 3.086 & 0.02 & 32 & 2.6 & -0.2500 & 0.23\\ 
\hline
\end{tabular}
\end{center}
\caption{\textbf{Free energy parameters used in the simulations.} Bulk BPI
correspond to Fig.~1 (\textbf{a}--\textbf{d}), bulk quench to
Fig.~1 (\textbf{e}--\textbf{h}); confined
geometries are shown in Fig.~2; and the external field case is
relevant for Fig.~3.}
\label{table:params}
\end{table}

{\bf Supplementary Methods}

@@

The surface anchoring for colloids is as before: always normal, but with
$WR/K = 0.23$ and $WR/K = 2.3$. The colloid solid volume fractions are
1\% and 4\%.
For both normal and planar anchoring at the walls,
the strength is adjusted to correspond to that used for the strongly
anchoring particles.
All simulations are run for two million simulation time steps.

\noindent
\textbf{External electric field.}
For the simulations of BPIII with colloidal particles a system size of
$128^3$ is used with a cholesteric pitch of $p=32$ in lattice units. 
The simulations are
performed at chirality $\kappa=2.6$ and reduced temperature $\tau=-0.25$.  
An initial configuration of
randomly oriented and positioned double twist cylinders is used,
from which an amorphous BPIII network emerges \cite{oliver-bp3}. This structure
is equilibrated for $6\times10^{5}$~LB time steps until no significant
further evolution is observed.

@@

\noindent
\textbf{Disclination rendering.}
In all cases visualisation of the defect lines is carried out via
an isosurface of the scalar order parameter~$q$ determined from the
largest eigenvalue of the local order parameter tensor. A low value
of the scalar order parameter~$q$ (typically in the range~0.12--0.14)
unambiguously identifies the disclinations.

@@

The timescale calibration may be obtained from (see e.g.~\cite{denniston})
\begin{equation}
\gamma_1=\frac{2q^2}{\mathit{\Gamma}},
\end{equation}
which relates the rotational viscosity $\gamma_1$ to the ordering strength
$q$ and the order parameter mobility $\mathit{\Gamma}$.
In the present simulations, $\mathit{\Gamma} = 0.5$ and the value of $\gamma$
for the simulation in Fig.~1 leads to $q=1/2$.   
For real liquid crystalline materials, $\gamma_1$ usually lies in range
$10^{-2}-1$ Pa s~\cite{deGennes}; for definiteness say
$\gamma_1 = 1$~Pa~s (equivalently 10 poise). Given the previous mapping for
free energy (or stress) units, this leads to a simulation unit of time equal
to $10^{-8}$~s. The total simulated time in Figs. 1-3 is therefore in the
millisecond range.

To calibrate the electric field strength 
one can equate the value of the dimensionless parameter ${\cal E}$ in the 
simulations with that of a hypothetical experiment.
Given the previous value for $A_0$, and a dielectric anisotropy of 20
(or equivalently $\epsilon_\mathrm{a}=20\epsilon_0$, with $\epsilon_0$ the
dielectric permittivity of vacuum), an electric field
of 100 V~$\mu$m$^{-1}$ is obtained. This corresponds to a value of
${\cal E}\sim 0.4$ (considering a value of $\gamma=3.0857$ as in
Fig.~3 of the main text).

@@

To summarise, the simulations represent BP-forming materials, with the
interpretation of the simulation units for length, time, and energy
density being close to 10~nm, 10~ns, and 100~MPa respectively.
For a Frank eleastic
constant of $K = 70$~pN and unit cell sizes in the range
$\lambda =$100--500~nm, a corresponding shear modulus $G \simeq
K/\lambda^2 =$ 0.3--7~kPa.
